Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-20 article.tex
typos

Considering the short delay of communication between neurons and the slow
decay of fluorescence, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in third plot of the figure:
\begin{align} % an asymmetrical discrete derivative filter
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i. \label{eq:high-pass-filter}
\end{align}

@@

Under that assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration or
precision matrix. Computationally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

To highlight the benefits of measuring the partial dependence conditional to
all other variables, we compare in Section~\ref{sec:results} with
\textit{Pearson correlation} statistics, which measure instead the linear dependence
between variables, independently of the others. Accordingly, $X_i$ and $X_j$ are
independent if their Pearson correlation statistic is zero, that is if
\begin{equation}
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0.
\end{equation}

Finally, let us remark that performance shows to be quite sensitive the choice
of hyper-parameters in our signal processing pipeline (e.g., $f_1$ or $f_2$ or
the value of the threshold $\tau$). To further improve robustness, the last
step of our method therefore consists in averaging correlation statistics over
various values of the parameters, thereby reducing the variance of our
predictions. Further details are provided in Appendix~\ref{app:optimized}.

>>>>

Considering the short delay of communication between neurons and the slow
decay of fluorescence, the second step of our pipeline consists in identifying
neuron spikes -- characterized by high frequencies -- by filtering out low
frequencies through a high-pass filter, resulting in the third plot of the figure:
\begin{align} % an asymmetrical discrete derivative filter
g(x^{t}_{i}) &= x^{t}_i - x^{t-1}_i. \label{eq:high-pass-filter}
\end{align}

@@

Under the assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration or
precision matrix. Computationally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

To highlight the benefits of measuring the partial dependence conditional to
all other variables, we compare in Section~\ref{sec:results} with the
\textit{Pearson correlation} statistics, which measure instead the linear dependence
between variables, independently of the others. Accordingly, $X_i$ and $X_j$ are
independent if their Pearson correlation statistic is zero, that is if
\begin{equation}
\rho_{X_i,X_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0.
\end{equation}

Finally, let us remark that performance shows to be quite sensitive to the choice
of hyper-parameters in the signal processing pipeline (e.g., $f_1$ or $f_2$ or
the value of the threshold $\tau$). To further improve robustness, the last
step of our method therefore consists in averaging correlation statistics over
various values of the parameters, thereby reducing the variance of our
predictions. Further details are provided in Appendix~\ref{app:optimized}.
